refnx - Neutron and X-ray reflectometry analysis in Python

### 2018-07-17 paper/manuscript.tex
PAPER: address comments

\newcommand{\refnx}{\emph{refnx }}
\newcommand{\Objective}{\textbf{Objective }}
\newcommand{\Parameter}{\textbf{Parameter }}
\newcommand{\Structure}{\textbf{Structure }}
\newcommand{\Slab}{\textbf{Slab }}
\newcommand{\Component}{\textbf{Component }}
\newcommand{\LipidLeaflet}{\textbf{LipidLeaflet }}
\newcommand{\Spline}{\textbf{Spline }}
\newcommand{\conda}{\emph{conda }}
\newcommand{\MixedReflectModel}{\emph{MixedReflectModel }}
\newcommand{\pip}{\emph{pip }}
\newcommand{\NumPy}{\emph{NumPy }}
\newcommand{\Cython}{\emph{Cython }}
\newcommand{\Jupyter}{\emph{Jupyter }}
\newcommand{\ipywidgets}{\emph{ipywidgets }}

@@

\begin{abstract}
The \refnx package is designed for object-oriented model-based analysis of neutron and X-ray reflectometry data in Python. It is cross platform, and has been tested on Linux, macOS, and Windows.
Structure construction is modular, being composed from a series of Components that each describe a subset of the interface. These Components (such as LipidLeaflet, or Spline) allow the user to parameterise their model in terms of physically relevant parameters. The Model and data are used to create an Objective, which is used to calculate residuals, log-likelihood, and log-prior probabilities of the system. Objectives are combined to perform co-refinement of multiple datasets, and mixed-area models. All parameters in the system can have a variety of Bounds applied; these bounds consist of probability distributions which represent their prior probability. The prior probability encodes a user's knowledge about the Parameter values, which allows the curve-fitting engine to direct solutions toward a more probable parameter space. Extra prior probability terms can be defined for Components, over and above those available from the Parameters alone.
Algebraic Parameter constraints are available.
The powerful curve-fitting engine offers the choice of using a normal least-squares fitting approach (with a variety of global and gradient-based optimizers) or a Bayesian approach using Markov Chain Monte Carlo to investigate the posterior distribution of the model parameters. The Bayesian approach is useful in examining parameter covariances, model selection, and variability in the resulting scattering length density profiles.

@@

Given that there are more and more publications arising from the reflectometry technique, it is vital that the analysis is reproducible. Reproducibility in research is an underlying principle of science. Unfortunately, it is not always possible to reproduce the results of others \cite{Stark2018}, because there is frequently not enough information provided in journal articles to repeat the analyses. Even if the datasets and software packages used to analyse them are supplied in supplementary information (most often they are not), a comprehensive set of instructions would need to be provided.
One example for addressing this reproducibility issue are the guidelines in the small-angle scattering community for the deposition of data and and associated models \cite{Trewhella:jc5010}.

Here we outline a new reflectometry analysis package, \refnx (version number x.x is used in this paper \cite{refnx}), that can address the reproducibility issue for the reflectometry community as it is specifically designed for use in \Jupyter notebooks \cite{Kluyver:2016aa} using a Python kernel \footnote{We do not mean that other programs are irreproducible, rather that the information provided by scientists in journal articles is often lacking.}.
These notebooks are a mix of executable code cells and markup. An analysis performed in such a notebook can be appended as supplementary information, along with the data, to an article. Readers can then replicate the exact data analysis, provided they have set up the same computing environment. Setting up the computing environment is simplified using the \conda package manager \cite{conda}, and an environment file (although other approaches are available). To that end the \Jupyter notebooks for the example analysis performed in this paper are appended as supplementary information.

@@

\subsection*{Method}
\refnx is written in Python with an object-oriented design, Figure~\ref{fig:components}. As with \emph{Motofit} \cite{Nelson2006} it calculates reflectivity using the Abeles method \cite{Heavens1955} for specular reflection from a stratified medium.
The building block of the analysis is the \Parameter object which represents a model value, whether it is allowed to vary in a fit, and a bounds attribute. The bounds are a probability distribution - the simplest bound being a uniform distribution which would set lower and upper limits for the parameter. This probability distribution is the prior probability for the parameter. Any of the \emph{scipy.stats} \cite{Jones2001-2017} continuous distributions, or distributions created by the user, can be used for this purpose. Algebraic relationships between \Parameter can be applied using constraints.
The \Structure object represents the interfacial model, comprising of individual \Component objects in series. Each \Component represents a subset of the interface and contains several physically relevant Parameters. The simplest \Component is a \Slab, which has a uniform scattering length density (SLD), thickness, roughness, and volume fraction of solvent. The simplest models are simply a series of \Slab. More sophisticated components include \LipidLeaflet and \Spline. \Spline is designed for freeform modelling of an SLD profile using spline interpolation. In addition to its constituent parameters each \Component can also contribute to the log-prior probability. This is useful when a \Component has a derived value, such as surface coverage, which is already known, and can be included as prior knowledge of the system. Each \Component has a \emph{slabs} property that represents the strata for its particular subset of the interface. A \Slab object has a single slice because it is a single thickness of uniform SLD. A \LipidLeaflet can be made of two slices (head/tail regions), but the \Spline Component has many. These slices have uniform SLD, and use the Nevot-Croce \cite{Nevot1980} to describe roughness between them. If a different roughness model is required then a 'microsliced' \Component could be written to accomodate that.
The \Structure object is used to construct a \textbf{ReflectModel} object. This object is responsible for calculating the resolution smeared reflectivity of the \Structure, scaling the data, and adding a linear background (via the scale and background Parameters). There are different types of smearing available - constant dQ/Q, point by point resolution smearing from the dataset of interest, or via a smearing kernel of arbitrary shape \cite{Nelson2014}.
The \Objective class uses the \textbf{ReflectModel} and a dataset \textbf{Data1D} to calculate $\chi^2$, log-likelihood, log-prior, residuals, and the generative model. The \Objective can be given a \textbf{Transform} object to permit fitting as log-R vs Q, $RQ^4$ vs Q, or lin-R vs Q. Several \Objective can be combined to form a \textbf{GlobalObjective} for co-refinement. The object-oriented nature allows \Parameter, \Component to be reused throughout the setup, thereby facilitating co-refinement. For an example please see \textbf{lipid.ipynb} in the supporting information.
The \Objective statistics are used directly by the \textbf{CurveFitter} class to perform least-square fitting or Bayesian MCMC of the system. \refnx uses the \emph{scipy} package for least-squares analysis (DifferentialEvolution, Levenberg Marquardt, LBFGSB - Limited Broyden–Fletcher–Goldfarb–Shanno with bounds) and the  \emph{emcee} package \cite{emcee} for affine invariant MCMC ensemble sampling of the posterior distribution of the parameters. The likelihoods used in MCMC and least-squares assume that the measurement errors are normally distributed, Equation~\ref{eqn:2}. However, other types of measurement errors (e.g. Poissonian) could be implemented by a subclass of \Objective and overriding the log-likelihood method.
Parallelisation of the sampling is done automatically, and can use MPI on a cluster for greater parallelisation. Visualisation of the samples produced by MCMC sampling is done using the \emph{corner} package \cite{corner} for scatter plot matrices. On the conclusion of a fit/sampling run each \Parameter is updated and given a standard error. For the sampling these represent the median and half the [15.87, 84.13] percentile range respectively; the latter approximates the standard deviation for a normally distributed statistic.

@@

To make it easy to create quick models, and to ease the learning curve, an ipywidgets \cite{ipywidgets} based fitting GUI can be created in a \Jupyter notebook, Figure~\ref{fig:gui}. This GUI builds slab based models for analysing a single dataset, but can be used as a stepping point for building more advanced models via the 'To code' code generation button.

>>>>

\newcommand{\refnx}{\emph{refnx }}
\newcommand{\Objective}{\textbf{Objective }}
\newcommand{\Parameter}{\textbf{Parameter }}
\newcommand{\Structure}{\textbf{Structure }}
\newcommand{\Slab}{\textbf{Slab }}
\newcommand{\Component}{\textbf{Component }}
\newcommand{\LipidLeaflet}{\textbf{LipidLeaflet }}
\newcommand{\DataD}{\textbf{Data1D }}
\newcommand{\ReflectModel}{\textbf{ReflectModel }}
\newcommand{\Spline}{\textbf{Spline }}
\newcommand{\conda}{\emph{conda }}
\newcommand{\MixedReflectModel}{\emph{MixedReflectModel }}
\newcommand{\pip}{\emph{pip }}
\newcommand{\NumPy}{\emph{NumPy }}
\newcommand{\Cython}{\emph{Cython }}
\newcommand{\Jupyter}{\emph{Jupyter }}
\newcommand{\ipywidgets}{\emph{ipywidgets }}

@@

\begin{abstract}
The \refnx package is designed for object-oriented model-based analysis of neutron and X-ray reflectometry data in Python. It is cross platform, and has been tested on Linux, macOS, and Windows.

Structure construction is modular, being composed from a series of Components that each describe a subset of the interface. These Components (such as LipidLeaflet, or Spline) allow the user to parameterise their model in terms of physically relevant parameters. The Model and data are used to create an Objective, which is used to calculate residuals, log-likelihood, and log-prior probabilities of the system. Objectives are combined to perform co-refinement of multiple datasets, and mixed-area models. All parameters in the system can have a variety of Bounds applied; these Bounds consist of probability distributions which represent their prior probability. The prior probability encodes a user's knowledge about the Parameter values, which allows the curve-fitting engine to direct solutions toward a more probable parameter space. Extra prior probability terms can be defined for Components, over and above those available from the Parameters alone.
Algebraic Parameter constraints are available.

The powerful curve-fitting engine offers the choice of using a normal least-squares fitting approach (with a variety of global and gradient-based optimizers) or a Bayesian approach using Markov Chain Monte Carlo to investigate the posterior distribution of the model parameters. The Bayesian approach is useful in examining parameter covariances, model selection, and variability in the resulting scattering length density profiles.

@@

Given that there are more and more publications arising from the reflectometry technique, it is vital that the analysis is reproducible. Reproducibility in research is an underlying principle of science. Unfortunately, it is not always possible to reproduce the results of others \cite{Stark2018}, because there is frequently not enough information provided in journal articles to repeat the analyses. Even if the datasets and software packages used to analyse them are supplied in supplementary information (most often they are not), a comprehensive, ordered, set of instructions would need to be provided.
One example for addressing this reproducibility issue are the guidelines in the small-angle scattering community for the deposition of data and and associated models \cite{Trewhella:jc5010}.

Here we outline a new reflectometry analysis package, \refnx (version number 0.1 is used in this paper \cite{refnx}), that can address the reproducibility issue for the reflectometry community as it is specifically designed for use in \Jupyter notebooks \cite{Kluyver:2016aa} using a Python kernel \footnote{We do not mean that other programs are irreproducible, rather that the information provided by scientists in journal articles is often lacking.}.
These notebooks are a mix of executable code cells and markup. An analysis performed in such a notebook can be appended as supplementary information, along with the data, to an article. Readers can then replicate the exact data analysis, provided they have set up the same computing environment. Setting up the computing environment is simplified using the \conda package manager \cite{conda}, and an environment file (although other approaches are available). To that end the \Jupyter notebooks for the example analysis performed in this paper are appended as supplementary information.

@@

\subsection*{Method}
\refnx is written in Python with an object-oriented design, Figure~\ref{fig:components}. As with \emph{Motofit} \cite{Nelson2006} it calculates reflectivity using the Abeles method \cite{Heavens1955} for specular reflection from a stratified medium.
The building block of the analysis is the \Parameter object which represents a model value, whether it is allowed to vary in a fit, and a bounds attribute. The bounds are a probability distribution - the simplest bound being a uniform distribution which would set lower and upper limits for the parameter. This probability distribution is the prior probability for the parameter. Any of the \emph{scipy.stats} \cite{Jones2001-2017} continuous distributions, or distributions created by the user, can be used for this purpose. Algebraic relationships between \Parameter can be applied using constraints.

The \Structure object represents the interfacial model, comprising of individual \Component objects in series. Each \Component represents a subset of the interface and contains several physically relevant Parameters. The simplest \Component is a \Slab, which has a uniform scattering length density (SLD), thickness, roughness, and volume fraction of solvent. The simplest models are simply a series of \Slab. More sophisticated components include \LipidLeaflet and \Spline. \Spline is designed for freeform modelling of an SLD profile using spline interpolation. In addition to its constituent parameters each \Component can also contribute to the log-prior probability. This is useful when a \Component has a derived value, such as surface coverage, which is already known, and can be included as prior knowledge of the system. Each \Component has a \emph{slabs} property that represents the strata for its particular subset of the interface. A \Slab object has a single slice because it is a single thickness of uniform SLD. A \LipidLeaflet can be made of two slices (head/tail regions), but the \Spline Component has many. These slices have uniform SLD, and use the Nevot-Croce approach \cite{Nevot1980} to describe roughness between them. If a different roughness model is required then a 'microsliced' \Component could be written to accomodate that.

The \Structure object is used to construct a \ReflectModel object. This object is responsible for calculating the resolution smeared reflectivity of the \Structure, scaling the data, and adding a linear background (via the scale and background Parameters). There are different types of smearing available - constant dQ/Q, point-by-point resolution smearing read from the dataset of interest, or via a smearing probability kernel of arbitrary shape \cite{Nelson2014}. The constant $dQ/Q$ and point-by-point smearing use Gaussian convolution, with $dQ$ representing the FWHM of a Gaussian approximation to the instrument resolution function \cite{Well2005}.

The \Objective class uses the \ReflectModel and a dataset, \DataD, to calculate $\chi^2$, log-likelihood, log-prior, residuals, and the generative model.
The \DataD object has \emph{x, y, y\_err, x\_err} attributes to represent Q, R, dR, dQ. The \DataD object can be constructed from a tuple of arrays representing the data, or by reading a three or four column plain-text datafile. A three column dataset represents Q ($\AA^{-1}$), R, dR (1 standard deviation). A four column dataset represents Q, R, dR, dQ ($\AA^{-1}$). Subclassing \DataD would allow other formats to be read. One example of this could be an wavelength dispersive file using $\Omega, \lambda$ data instead of Q; such as that used in energy scanned X-ray reflectometry, or sometimes produced by wavelength dispersive neutron reflectometers in order to account for absorption terms. In such a case \ReflectModel could be subclassed to make full use of this information.
The \Objective can be given a \textbf{Transform} object to permit fitting as log-R vs Q, $RQ^4$ vs Q, or lin-R vs Q. Several \Objective can be combined to form a \textbf{GlobalObjective} for co-refinement. The object-oriented nature allows \Parameter, \Component to be reused throughout the setup, thereby facilitating co-refinement.
For an example please see \textbf{lipid.ipynb} in the supporting information.

The \Objective statistics are used directly by the \textbf{CurveFitter} class to perform least-square fitting or Bayesian MCMC of the system. \refnx uses the \emph{scipy} package for least-squares analysis (DifferentialEvolution, Levenberg Marquardt, LBFGSB - Limited Broyden–Fletcher–Goldfarb–Shanno with bounds) and the  \emph{emcee} package \cite{emcee} for affine invariant MCMC ensemble sampling of the posterior distribution of the parameters. The likelihoods used in MCMC and least-squares assume that the measurement errors are normally distributed, Equation~\ref{eqn:2}. However, other types of measurement errors (e.g. Poissonian) could be implemented by a subclass of \Objective  overriding the log-likelihood method.
Parallelisation of the sampling is done automatically, and can use MPI on a cluster for greater parallelisation. Visualisation of the samples produced by MCMC sampling is done using the \emph{corner} package \cite{corner} for scatter plot matrices. On the conclusion of a fit/sampling run each \Parameter is updated and given a standard error. For the sampling these represent the median and half the [15.87, 84.13] percentile range respectively; the latter approximates the standard deviation for a normally distributed statistic.

@@

To make it easy to create quick models, and to ease the learning curve, an ipywidgets \cite{ipywidgets} based fitting GUI can be created in a \Jupyter notebook, Figure~\ref{fig:gui}. This GUI builds slab based models for analysing a single dataset, but can be used as a stepping point for building more advanced models via the 'To code' code generation button. A fully functional web-based reflectometry analysis notebook is currently available \cite{Nelson2018}.
